{\steve}: A hierarchical Bayesian model for Supernova Cosmology

### 2018-10-15 papers/methods/paper.tex
Some more updates from ChrisL

The benefit this analysis methodology provides is speed -- for samples of hundreds of supernovae or less, efficient matrix inversion algorithms allow the likelihood to be evaluated quickly. The speed comes with several limitations. Firstly, formulating a $\chi^2$ likelihood results in a loss of model flexibility by assuming Gaussian uncertainty. Secondly, the method of creating a covariance matrix relies on computing partial derivatives and thus any uncertainty estimated from this method loses all information about correlation between sources of uncertainty. For example, the underlying supernova color population's mean and skewness are highly correlated, however this correlation is ignored when determining population uncertainty using numerical derivatives of population permutations. Whilst correlations can be incorporated into a covariance matrix, it requires human awareness of the correlations and thus methods which can automatically capture correlated uncertainties are preferable. Thirdly, the computational efficiency is dependent on both creating the global covariance matrix, and then inverting a covariance matrix with dimensionality linearly proportional to the number of supernovae. As this number increases, the cost of inversion rises quickly, and is not viable for samples with thousands of supernovae. A recent solution to this computational cost problem is to bin the supernovae in redshift bins, which produces a matrix of manageable size and can allow fitting without matrix inversion on every step. Whilst binning data results in some loss of information, recent works tested against simulations show that this loss of information does not create significant cosmological biases \citep{Scolnic2016, Kessler2017}.

@@

Simulations provide great flexibility and freedom in how to treat the systematic uncertainties and selection effects associated with supernova surveys. By using forward modelling directly from these simulations, data does not need to be corrected, analytic approximations do not need to be applied; we are free to incorporate algorithms that simply cannot be expressed in a tractable likelihood such as those found in traditional analyses from Section \ref{sec:traditional}. This freedom comes with a cost -- computation. The classical $\chi^2$ method's most computationally expensive step in a fit is matrix inversion. For ABC methods, we must instead simulate an entire supernova population -- drawing from underlying supernova populations, modelling light curves, applying selection effects, fitting light curves and applying data cuts. This is an intensive process.

@@

With the introduction of multiple layers in our model, we can add more flexibility than a traditional analysis whilst still maintaining most of the computational benefits that come from having a tractable likelihood. \citet{Mandel2009, Mandel2011a, Mandel2017} construct a hierarchical model that they apply to supernova light-curve fitting. \citet{March2011} derive a hierarchical model and simplify it by analytically marginalising over nuisance parameters to provide increased flexibility with reduced uncertainty over the traditional method, but do not incorporate bias correction. \citet{March2014} and \citet{Karpenka2015} improve upon this by incorporating redshift-dependent magnitude corrections from \citet{Perrett2010} to remove bias, and validate on 100 realisations of SNLS-like simulations. The recent BAHAMAS model \citep{Shariff2016} builds on this and reanalyses the JLA dataset \citep[using redshift dependent bias corrections from][]{Betoule2014}, whilst including extra freedom in the correction factors $\alpha$ and $\beta$, finding evidence for redshift dependence on $\beta$. \citet{Ma2016} performed a reanalysis of the JLA dataset within a Bayesian formulation, finding significant differences in $\alpha$ and $\beta$ values from the original analysis from \citet{Betoule2014}. Notably, these methods rely on data that is bias corrected or the methods ignore biases, however the UNITY framework given by \citet{Rubin2015} incorporates selection efficiency analytically in the model, and is applied to the Union 2.1 dataset \citep{Suzuki2012}. The assumption made by the UNITY analysis is that the bias in real data is perfectly described by an analytic function. They validate their  model to be free of significant biases using fits to thirty realisations of supernova datasets that are constructed to mimic the UNITY framework. The well known BEAMS (Bayesian estimation applied to multiple species) methodology from \citet{Kunz2007} has been extended and applied in several works \citep{Hlozek2012}, most lately to include redshift uncertainty for photometric redshift application as zBEAMS \citep{Roberts2017} and to include simulated bias corrections in \citet{Kessler2017}. For the latter case, by inferring biases using Bayesian models, sophisticated corrections can be calculated and then applied to more traditional $\chi^2$ models.

Whilst there are a large number of hierarchical models available, none of them have undergone high-statistic tests using realistic simulations to verify each models' respective bias. Additionally, testing has generally been performed on supernovae simulations with either $\Lambda$CDM cosmology or Flat $\Lambda$CDM cosmology. However, quantifying the biases on $w$CDM cosmology simulations with realistic simulations is becoming critically important as precision supernovae cosmology comes into its own, and focus shifts from determination of $\Omega_m$ to both $\Omega_m$ and $w$.

@@

Finally, one of the other persistent challenges facing supernova cosmology analyses are the high number of systematics. Because of the rarity of SN~Ia events, datasets are commonly formed from the SN~Ia discoveries of multiple surveys in order to increase the number of supernovae in a dataset. However, each different survey introduces additional sources of systematic error, from sources within each survey such as band calibration, to systematics introduced by calibration across surveys. Peculiar velocities, different host environments, and dust extinction represent additional sources of systematic uncertainty which must all be modelled and accounted for.

@@

here are numerous results showing statistically significant correlations between host-galaxy environment and supernova properties \citep{Kelly2010, Lampeitl2010, Sullivan2010, DAndrea2011, Gupta2011, Johansson2013, Rigault2013}. The latest sample of over 1300 spectroscopically confirmed type Ia supernovae show $>5\sigma$ evidence for correlation between host mass and luminosity \citep{Uddin2017}. The traditional correction, as employed in analyses such as \citet{Suzuki2012} and \citet{Betoule2014}, invokes a step function such that $\Delta M = \gamma \mathcal{H}(\log(M) - 10))$, where $\mathcal{H}$ is the Heaviside step function, $M$ is the galaxy mass in solar masses and $\gamma$ represents the size of the magnitude step. The scale of this step function varies from analysis to analysis, and is treated as a fit parameter. In this work we adopt the model used in {\rubin}, which follows the work from \citet{Rigault2013}, such that we introduce two parameters to incorporate a redshift-dependent host galaxy mass correction:

@@

\begin{equation}
P\left(\hat{\eta}, \frac{\partial \hat{\eta}}{\partial \Z_i} | \eta, \delta \Z_i, C\right) = \mathcal{N}\left(\hat{\eta} + \delta \Z_i \frac{\partial \hat{\eta}}{\partial \Z_i}|\eta,\cov\right). \label{eq:l3}
\end{equation}

@@

\begin{figure}
	\begin{center}
		\includegraphics[width=\columnwidth]{shift.pdf}
	\end{center}
	\caption{Testing the correctness of our normal approximation to the skewed color distribution. The `correct' line (shown in black) represents the exact integral $w = \int P(S|x) P(x) dx$ where $P(S|x)$ is an error function (following our high-redshift surveys) and $P(x) = \mathcal{N}^{\rm Skew}(x, 0.1, 2)$, calculated numerically. The $x$-axis is analogous to $m_B$ is cosmological context. As expected, all efficiencies drop towards zero as shift increases (as objects get fainter). The unshifted normal approximation shows significant discrepancy in the calculated efficiency as it transitions from 1 to 0, whilst the shifted normal approximation shows negligible error to the correct solution. From these plots, further refinement of the normal approximation (such as including kurtosis or higher powers) as unnecessary.}
	\label{fig:shift}
\end{figure}

@@

The selection functions parameters (a skew normal for low-redshift and an error function for high-redshift) are all given independent uncertainty of $0.01$ (mean and width for the CDF selection function, and mean, width and skewness for the skew normal selection function). Draw from each survey simulation until we have 1000 low-$z$ supernovae and 1000 DES-like supernovae, representing a statistical sample of greater power than the estimated 350 supernovae for the DES-SN3YR sample. Sample data for 1000 high and low redshift supernovae are shown in Figure \ref{fig:simple_pop}, confirming the presence of strong selection effects in both toy surveys, as designed.

@@

\subsection{General Selection Effects}
\label{app:selection1}
When formulating and fitting a model using a constraining dataset, we wish to resolve the posterior surface defined by
\begin{equation}
P(\theta | {\rm data}) \propto P({\rm data} | \theta) P(\theta),
\end{equation}
which gives the probability of the model parameter values ($\theta$) given the data.  Prior knowledge of the allowed values of the model parameters is encapsulated in the prior probability $P(\theta)$. Of primary interest to us is the likelihood of observing the data given our parametrised model, $\mathcal{L} \equiv P({\rm data} | \theta)$. When dealing with experiments that have imperfect selection efficiency, our likelihood must take that efficiency into account.  We need to describe the probability that the events we observe are both drawn from the distribution predicted by the underlying theoretical model \textit{and} that those events, given they happened, are subsequently successfully observed.  To make this extra conditional explicit, we write the likelihood of the data given an underlying model, $\theta$, \textit{and} that the data are included in our sample, denoted by $S$, as:
\begin{equation}
\mathcal{L} = P({\rm data} | \theta, S). \label{eq:app_like}
\end{equation}
A variety of selection criteria are possible, and in our method we use our data in combination with the proposed model to determine the probability of particular selection criteria.  That is, we characterise a function $P(S|{\rm data},\theta)$, which colloquially can be stated as \textit{the probability of a potential observation passing selection cuts, given our observations and the underlying model}. We can introduce this expression in a few lines due to symmetry of joint probabilities and utilising that $P(x,y,z) = P(x|y,z)P(y,z) = P(y|x, z)P(x, z)$:
\begin{eqnarray}
P({\rm data} | S , \theta) P(S, \theta) &=& P(S | {\rm data}, \theta) P({\rm data}, \theta)\\
P({\rm data} | S , \theta) &=& \frac{ P(S | {\rm data}, \theta) P({\rm data}, \theta) }{ P(S, \theta) }\\
&=& \frac{ P(S | {\rm data}, \theta) P({\rm data} | \theta) P(\theta)}{ P(S | \theta)  P(\theta)}\\
&=& \frac{ P(S | {\rm data}, \theta) P({\rm data} | \theta)}{ P(S | \theta) }
\end{eqnarray}
which is equal to the likelihood $\mathcal{L}$. Introducing an integral over all possible events $D$, so we can evaluate $P(S|\theta)$, 
\begin{eqnarray}
\mathcal{L}&=& \frac{P(S|{\rm data},\theta) P({\rm data}|\theta)}{\int P(S, D|\theta)\, dD} \\
\mathcal{L} &=& \frac{P(S|{\rm data},\theta) P({\rm data}|\theta)}{\int P(S | D, \theta) P(D|\theta)\, dD}, \label{eq:app_main}
\end{eqnarray}
where we define the denominator as $d$ for simplicity in future derivations.

@@

\begin{figure*}
	\begin{center}
		\includegraphics[width=\textwidth]{simple_w_shift_dist_0.pdf}
	\end{center}
	\caption{Marginalised probability distributions for 100 realisations of cosmology, fit to Flat $w$CDM with prior $\Omega_m \sim \mathcal{N}(0.3, 0.01)$, each containing 1000 simulated high-$z$ and 1000 simulated low-$z$ supernovae. The dashed green surfaces represent a fit to an underlying Gaussian color population with the unshifted model. The blue solid surface represents fits to a skewed color population with the unshifted model, and the purple dotted surface represents a fit to a skewed color population with the shifted model. The superscript $0$ and $1$ denote the two different surveys (high-$z$ and low-$z$ respectively), and similarly the first four $\langle c^i \rangle$ parameters represent the four redshift nodes in the high-$z$ survey, and the last four represent the nodes for the low-$z$ survey. We can see that the shifted model is far better able to recover skewed input populations than the unshifted, performing better in terms of recovering skewness $\alpha_c$, mean color $\langle c \rangle$ and width of the color distribution $\sigma_c$. The unshifted model recovers the correct color mean and width if you approximate a skew normal as a normal: $\Delta\mu = \sqrt{2/\pi}\sigma_c\delta_c \approx 0.071$, which is approximately the deviation found in fits to the color population mean. Importantly, the unshifted model when run on skewed data (the solid blue) shows extreme bias in $\alpha_c$, where it fits strongly around zero regardless, showing it to be a poor approximation. Based on these results and the fantastic performance in correctly recovering underlying populations of the shifted normal approximation, we adopt the shifted normal approximation in our model.}
	\label{fig:simple_w_super}
\end{figure*}

@@

The simplest and most commonly found optimisation we employed was to precompute as much as possible to reduce the complexity of the mathematical graph our model is translated into to compute the surface derivatives. For example, when computing the distance modulus, redshift is encountered to various powers. Instead of computing those powers in Stan, we simply pass in several arrays of redshift values already raised to the correct power. Small changes like this however only give small improvements.

The primary numerical improvement we made on existing frameworks was to remove costly probability evaluations of multivariate normals. To increase efficiency, the optimum way to sample a multivariate normal is to reparameterise it such that instead of sampling $\mathcal{N}(\vec{x}|\vec{\mu}, \Sigma)$, you sample $\mathcal{N}(\vec{\delta}|0,1)$ where $\vec{x} = \vec{\mu} + L \vec{\delta}$ and $L$ is the cholesky decomposition of $\Sigma$. In this way, we can efficiently sample the unit normal probability distribution instead of sampling a multivariate normal probability distribution. Switching to this parametrisation resulted in a computational increase of an order of magnitude, taking fits for a sample of approximately 500 supernovae from roughly four hours down to thirty minutes.

>>>>

The benefit this analysis methodology provides is speed -- for samples of hundreds of supernovae or less, efficient matrix inversion algorithms allow the likelihood to be evaluated quickly. The speed comes with several limitations. Firstly, formulating a $\chi^2$ likelihood results in a loss of model flexibility by assuming Gaussian uncertainty. Secondly, the method of creating a covariance matrix relies on computing partial derivatives and thus any uncertainty estimated from this method loses information about correlation between sources of uncertainty. For example, the underlying supernova color population's mean and skewness are highly correlated, however this correlation is ignored when determining population uncertainty using numerical derivatives of population permutations. Whilst correlations can be incorporated into a covariance matrix, it requires human awareness of the correlations and thus methods that can automatically capture correlated uncertainties are preferable. Thirdly, the computational efficiency is dependent on both creating the global covariance matrix, and then inverting a covariance matrix with dimensionality linearly proportional to the number of supernovae. As this number increases, the cost of inversion rises quickly, and is not viable for samples with thousands of supernovae. A recent solution to this computational cost problem is to bin the supernovae in redshift bins, which produces a matrix of manageable size and can allow fitting without matrix inversion at every step. Whilst binning data results in some loss of information, recent works tested against simulations show that this loss does not result in significant cosmological biases \citep{Scolnic2016, Kessler2017}.

@@

Simulations provide great flexibility and freedom in how to treat the systematic uncertainties and selection effects associated with supernova surveys. By using forward modelling directly from these simulations, data does not need to be corrected, analytic approximations do not need to be applied, and we are free to incorporate algorithms that simply cannot be expressed in a tractable likelihood such as those found in traditional analyses from Section \ref{sec:traditional}. This freedom comes with a cost -- computation. The classical $\chi^2$ method's most computationally expensive step in a fit is matrix inversion. For ABC methods, we must instead simulate an entire supernova population -- drawing from underlying supernova populations, modelling light curves, applying selection effects, fitting light curves and applying data cuts. This is an intensive process.

@@

With the introduction of multiple layers in our model, we can add more flexibility than a traditional analysis whilst still maintaining most of the computational benefits that come from having a tractable likelihood. \citet{Mandel2009, Mandel2011a, Mandel2017} construct a hierarchical model that they apply to supernova light-curve fitting. \citet{March2011} derive a hierarchical model and simplify it by analytically marginalising over nuisance parameters to provide increased flexibility with reduced uncertainty over the traditional method, but do not incorporate bias corrections. \citet{March2014} and \citet{Karpenka2015} improve upon this by incorporating redshift-dependent magnitude corrections from \citet{Perrett2010} to remove bias, and validate on 100 realisations of SNLS-like simulations. The recent BAHAMAS model \citep{Shariff2016} builds on this and reanalyses the JLA dataset \citep[using redshift dependent bias corrections from][]{Betoule2014}, whilst including extra freedom in the correction factors $\alpha$ and $\beta$, finding evidence for redshift dependence on $\beta$. \citet{Ma2016} performed a reanalysis of the JLA dataset within a Bayesian formulation, finding significant differences in $\alpha$ and $\beta$ values from the original analysis from \citet{Betoule2014}. Notably, these methods rely on data that is bias corrected or the methods ignore biases, however the UNITY framework given by \citet{Rubin2015} incorporates selection efficiency analytically in the model, and is applied to the Union 2.1 dataset \citep{Suzuki2012}. The assumption made by the UNITY analysis is that the bias in real data is perfectly described by an analytic function. They validate their  model to be free of significant biases using fits to thirty realisations of supernova datasets that are constructed to mimic the UNITY framework. The well known BEAMS (Bayesian estimation applied to multiple species) methodology from \citet{Kunz2007} has been extended and applied in several works \citep{Hlozek2012}, most lately to include redshift uncertainty for photometric redshift application as zBEAMS \citep{Roberts2017} and to include simulated bias corrections in \citet{Kessler2017}. For the latter case, by inferring biases using Bayesian models, sophisticated corrections can be calculated and then applied to more traditional $\chi^2$ models.

Whilst there are a large number of hierarchical models available, none of them have undergone comprehensive tests using realistic simulations to verify each models' respective bias. Additionally, testing has generally been performed on supernovae simulations with either $\Lambda$CDM cosmology or Flat $\Lambda$CDM cosmology. However, quantifying the biases on $w$CDM cosmology simulations with realistic simulations is becoming critically important as precision supernovae cosmology comes into its own, and focus shifts from determination of $\Omega_m$ to both $\Omega_m$ and $w$.

@@

Finally, one of the other persistent challenges facing supernova cosmology analyses are the high number of systematics. Because of the rarity of SN~Ia events, datasets are commonly formed from the SN~Ia discoveries of multiple surveys in order to increase the number of supernovae in a dataset. However, each survey introduces additional sources of systematic error, from sources within each survey such as band calibration, to systematics introduced by calibration across surveys. Peculiar velocities, different host environments, and dust extinction represent additional sources of systematic uncertainty which must all be modelled and accounted for.

@@

There are numerous results showing statistically significant correlations between host-galaxy environment and supernova properties \citep{Kelly2010, Lampeitl2010, Sullivan2010, DAndrea2011, Gupta2011, Johansson2013, Rigault2013}. The latest sample of over 1300 spectroscopically confirmed type Ia supernovae show $>5\sigma$ evidence for correlation between host mass and luminosity \citep{Uddin2017}. The traditional correction, as employed in analyses such as \citet{Suzuki2012} and \citet{Betoule2014}, invokes a step function such that $\Delta M = \gamma \mathcal{H}(\log(M) - 10))$, where $\mathcal{H}$ is the Heaviside step function, $M$ is the galaxy mass in solar masses and $\gamma$ represents the size of the magnitude step. The scale of this step function varies from analysis to analysis, and is treated as a fit parameter. In this work we adopt the model used in {\rubin}, which follows the work from \citet{Rigault2013}, such that we introduce two parameters to incorporate a redshift-dependent host galaxy mass correction:

@@

\begin{equation}
P\left(\hat{\eta}, \frac{\partial \hat{\eta}}{\partial \Z_i} | \eta, \delta \Z_i, C_\eta\right) = \mathcal{N}\left(\hat{\eta} + \delta \Z_i \frac{\partial \hat{\eta}}{\partial \Z_i}|\eta,\cov\right). \label{eq:l3}
\end{equation}

@@

\begin{figure}
	\begin{center}
		\includegraphics[width=\columnwidth]{shift.pdf}
	\end{center}
	\caption{Testing the correctness of our normal approximation to the skewed color distribution. The `correct' line (shown in black) represents the exact integral $w = \int P(S|x) P(x) dx$ where $P(S|x)$ is an error function (following our high-redshift surveys) and $P(x) = \mathcal{N}^{\rm Skew}(x, 0.1, 2)$, calculated numerically. The $x$-axis is analogous to $m_B$ is cosmological context. As expected, all efficiencies drop towards zero as shift increases (as objects get fainter). The unshifted normal approximation shows significant discrepancy in the calculated efficiency as it transitions from 1 to 0, whilst the shifted normal approximation shows negligible error to the correct solution. From these plots, further refinement of the normal approximation (such as including kurtosis or higher powers) are unnecessary.}
	\label{fig:shift}
\end{figure}

@@

The selection functions parameters (a skew normal for low-redshift and a complimentary error function for high-redshift) are all given independent uncertainty of $0.01$ (mean and width for the CDF selection function, and mean, width and skewness for the skew normal selection function). Draw from each survey simulation until we have 1000 low-$z$ supernovae and 1000 DES-like supernovae, representing a statistical sample of greater power than the estimated 350 supernovae for the DES-SN3YR sample. Sample data for 1000 high and low redshift supernovae are shown in Figure \ref{fig:simple_pop}, confirming the presence of strong selection effects in both toy surveys, as designed.

@@

\subsection{General Selection Effects}
\label{app:selection1}
When formulating and fitting a model using a constraining dataset, we wish to resolve the posterior surface defined by
\begin{equation}
P(\theta | {\rm data}) \propto P({\rm data} | \theta) P(\theta),
\end{equation}
which gives the probability of the model parameter values ($\theta$) given the data.  Prior knowledge of the allowed values of the model parameters is encapsulated in the prior probability $P(\theta)$. Of primary interest to us is the likelihood of observing the data given our parametrised model, $\mathcal{L} \equiv P({\rm data} | \theta)$. When dealing with experiments that have imperfect selection efficiency, our likelihood must take that efficiency into account.  We need to describe the probability that the events we observe are both drawn from the distribution predicted by the underlying theoretical model \textit{and} that those events, given they happened, are subsequently successfully observed.  To make this extra conditional explicit, we write the likelihood of the data given an underlying model, $\theta$, \textit{and} that the data are included in our sample, denoted by $S$, as:
\begin{equation}
\mathcal{L}(\theta; {\rm data}) = P({\rm data} | \theta, S). \label{eq:app_like}
\end{equation}
A variety of selection criteria are possible, and in our method we use our data in combination with the proposed model to determine the probability of particular selection criteria.  That is, we characterise a function $P(S|{\rm data},\theta)$, which colloquially can be stated as \textit{the probability of a potential observation passing selection cuts, given our observations and the underlying model}. We can introduce this expression in a few lines due to symmetry of joint probabilities and utilising that $P(x,y,z) = P(x|y,z)P(y,z) = P(y|x, z)P(x, z)$:
\begin{eqnarray}
P({\rm data} | S , \theta) P(S, \theta) &=& P(S | {\rm data}, \theta) P({\rm data}, \theta)\\
P({\rm data} | S , \theta) &=& \frac{ P(S | {\rm data}, \theta) P({\rm data}, \theta) }{ P(S, \theta) }\\
&=& \frac{ P(S | {\rm data}, \theta) P({\rm data} | \theta) P(\theta)}{ P(S | \theta)  P(\theta)}\\
&=& \frac{ P(S | {\rm data}, \theta) P({\rm data} | \theta)}{ P(S | \theta) }
\end{eqnarray}
which is equal to the likelihood $\mathcal{L}$. Introducing an integral over all possible events $D$, so we can evaluate $P(S|\theta)$, 
\begin{eqnarray}
\mathcal{L}(\theta; {\rm data})&=& \frac{P(S|{\rm data},\theta) P({\rm data}|\theta)}{\int P(S, D|\theta)\, dD} \\
\mathcal{L}(\theta; {\rm data}) &=& \frac{P(S|{\rm data},\theta) P({\rm data}|\theta)}{\int P(S | D, \theta) P(D|\theta)\, dD}, \label{eq:app_main}
\end{eqnarray}
where we define the denominator as $d$ for simplicity in future derivations.

@@

\begin{figure*}
	\begin{center}
		\includegraphics[width=\textwidth]{simple_w_shift_dist_0.pdf}
	\end{center}
	\caption{Marginalised probability distributions for 100 realisations of cosmology, fit to Flat $w$CDM with prior $\Omega_m \sim \mathcal{N}(0.3, 0.01)$, each containing 1000 simulated high-$z$ and 1000 simulated low-$z$ supernovae. The dashed green surfaces represent a fit to an underlying Gaussian color population with the unshifted model. The blue solid surface represents fits to a skewed color population with the unshifted model, and the purple dotted surface represents a fit to a skewed color population with the shifted model. The superscript $0$ and $1$ denote the two different surveys (high-$z$ and low-$z$ respectively), and similarly the first four $\langle c^i \rangle$ parameters represent the four redshift nodes in the high-$z$ survey, and the last four represent the nodes for the low-$z$ survey. We can see that the shifted model is far better able to recover skewed input populations than the unshifted, performing better in terms of recovering skewness $\alpha_c$, mean color $\langle c \rangle$ and width of the color distribution $\sigma_c$. The unshifted model recovers the correct color mean and width if you approximate a skew normal as a normal: $\Delta\mu = \sqrt{2/\pi}\sigma_c\delta_c \approx 0.071$, which is approximately the deviation found in fits to the color population mean. Importantly, the unshifted model when run on skewed data (the solid blue) shows extreme bias in $\alpha_c$, where it fits strongly around zero regardless, showing it to be a poor approximation. Based on these results and the good performance in correctly recovering underlying populations of the shifted normal approximation, we adopt the shifted normal approximation in our model.}
	\label{fig:simple_w_super}
\end{figure*}

@@

The simplest and most commonly found optimisation we employed was to precompute as much as possible. This is in a bid to reduce the complexity of the mathematical graph our model is translated into by Stan, to simplify the computation of surface derivatives. For example, when computing the distance modulus, redshift is encountered to various powers. Instead of computing those powers in Stan, we simply pass in several arrays of redshift values already raised to the correct power. Small changes like this however only give small improvements.

The primary numerical improvement we made on existing frameworks was to remove costly probability evaluations of multivariate normals. To increase efficiency, the optimum way to sample a multivariate normal is to reparameterise it such that instead of sampling $\mathcal{N}(\vec{x}|\vec{\mu}, \Sigma)$, we sample $\mathcal{N}(\vec{\delta}|0,1)$ where $\vec{x} = \vec{\mu} + L \vec{\delta}$ and $L$ is the cholesky decomposition of $\Sigma$. In this way, we can efficiently sample the unit normal probability distribution instead of sampling a multivariate normal probability distribution. Switching to this parametrisation resulted in a computational increase of an order of magnitude, taking fits for a sample of approximately 500 supernovae from roughly four hours down to thirty minutes.
